Type theories as quotient inductive-recursive types

### 2025-02-13 tex/main.tex
Draft S4.3

We work in intensional Martin-L\"of type theory with uniqueness of identity proofs. In \cref{sec:models,sec:formal-implementation}, we also assume function extensionality and $\eta$-rules for $\Pi$-types and $\Sigma$-types.
%
We mostly follow the notational conventions of the HoTT book~\cite{UFP2013} (even if we do not assume the univalence axiom), except that we write dependent function types using Agda notation $(x : A) \to B\,x$.
%
We are further inspired by Agda in writing $\implicit{x}\;B\,x$ for $(x : A) \to B\,x$ when $x$ is treated as an `implicit' argument which is not given explicitly, but instead can be figured out from context.
We write $\Set_i$ for the cumulative hierarchy of universes with the level index $i$ suppressed.

@@

Quotient inductive types (QITs) are a special case of higher inductive types~\cite{Lumsdaine2020} from homotopy type theory, where constructors can not only introduce new elements, but also new equalities between the introduced elements.
%
Since we assume uniqueness of identity proofs globally, all QITs are implicitly truncated to be sets in the sense of homotopy type theory.
%
For example, the type of \emph{join lists} $\JList{A}$ can be given by the QIT
with the following constructors:
\begin{alignat*}{3}
  \Jnil                    & : \implicit{A}      && \JList{A} \\
  \Jsing{\blank}           & : \implicit{A}      && A \to \JList{A} \\
  \Jconcat{\blank}{\blank} & : \implicit{A}      && \JList{A} \to \JList{A} \to \JList{A} \\
  \mathsf{idl}             & : \implicit{\ell}\; && (\Jconcat{\ell}{\Jnil}) =^{\JList{A}} \ell \\
  \mathsf{idr}             & : \implicit{\ell}\; && (\Jconcat{\Jnil}{\ell}) =^{\JList{A}} \ell \\
  \mathsf{assoc}           & : \implicit{\ell_0, \ell_1, \ell_2}\; && \Jconcat{(\Jconcat{\ell_0}{\ell_1})}{\ell_2} =^{\JList{A}} \Jconcat{\ell_0}{(\Jconcat{\ell_1}{\ell_2})}
\end{alignat*}

@@

\subsection{Definitions by overlapping patterns} \label{sec:meta:overlapping}
\begin{figure} 
\begin{minipage}[t]{.5\textwidth}
  \begin{alignat*}{5}
    \blank  & + \blank     && : \N \to \N \to \N \\
    0       &  + m         && = m  \\
  n         &  + 0         && = m  \\
  (\suc\;n) &  + m         && = \suc\;(n + m) \\
  n         &  + (\suc\;m) && = \suc\;(n + m)
\end{alignat*}
\end{minipage}
\begin{minipage}[t]{.5\textwidth}
\begin{alignat*}{5}
  \blank         & +'\blank     && : \N \to \N \to \N \\
  0              & +' m         && \defeq m  \\
  (\suc\;n)      & +' 0         && \defeq \suc\;n \\
  (\suc\;n)      & +' (\suc\;m) && \defeq \suc\;(n +' m)
\end{alignat*}
\end{minipage}
\caption{Two seemingly different definitions of addition}
\label{fig:additions}
\end{figure}
Functions are typically defined informally using pattern matching for convenience and readability. Formally, however, they are either defined using an explicit pattern-matching construct or elaborated into case trees, which correspond to the \emph{eliminator}, following \emph{first-match} semantics~\cite{Cockx2020a} and employing termination and completeness checking.
This elaboration enables conventional notation while making computational behaviour opaque and favouring asymmetric definitions.
For example, two definitions of addition in \cref{fig:additions} are the same
after elaboration in \Agda: eliminating the first argument of $+$ distinguishes $0$ and $\suc\;n$ and refines the second and the third clauses into clauses of $+'$ and makes the fourth clause useless.
Each clause of $+'$ is strict (hence the symbol $\defeq$ is used) while clauses of $+$ are not necessarily strict by definition.
This semantics for function definitions decides how definitions reduce and therefore how identities are proved.

Cockx~\cite{Cockx2014} explores an alternative semantics for pattern matching, where all clauses are strict, even in the presence of \emph{overlapping patterns}.
Thus, when a function $f$ is defined with overlapping patterns, the notation $f\;\overrightarrow{p_i} \reduce t_i$ is used to emphasize its reduction behaviour.
Under this semantics, each clause of $\blank + \blank$ is strictly equal to its right-hand side, allowing the symmetry addition to be proved commutative directly by induction on $n$ and $m$.

@@

The first and second projections of $(\sigma, t)$ should obviously be $\sigma$ and $t$ respectively
\begin{alignat*}{5}
  \pi_1\beta      & : \implicit{\Gamma, \Delta, \Theta, \sigma, A, t} && \pi_1(\sigma , t)        && =^{\Sub\,\Gamma,\Delta} &&  \sigma \\
  \pi_2\beta      & : \implicit{\Gamma, \Delta, \Theta, \sigma, A, t} && \pi_2(\sigma , t)        && =^{\Tm\,\Gamma\,([\blank]\,A)}_{\alert{\pi_1\beta}} &&  t, 
\end{alignat*}
but again, note that the type of $\pi_2\beta$ only makes sense because we can transport along $\pi_1\beta$ to make sure both sides of the equation have the same type: $\pi_2(\sigma, t)$ is an habitant of $\Tm\;\Gamma\;([\pi_1\,(\sigma, t)]\,A)$ instead of $\Tm\;\Gamma\;([\sigma]\,A)$.
%whereas $\pi_2(\sigma, t)$ is an habitant of $\Tm\;\Gamma\;([\pi_1\,(\sigma, t)]\,A)$ instead of $\Tm\;\Gamma\;([\sigma]\,A)$ on the right-hand side. 
%Therefore, we have to transport $\pi_2(\sigma, t)$ along $\pi_1\beta$, so the above equality constructor is, in fact, $\transfib{\Tm\;\Gamma\;([\blank]\,A)}{\pi_1\beta}{\pi_2(\sigma, t)} =^{\Tm\;\Gamma\;([\sigma]\,A)} t$.
%
Similarly, for term substitution, terms are transported along the corresponding rules:
\[
  [\idS]_{\Tm} : \implicit{\Gamma, A, t}\; {[\idS]\,t} =^{\Tm\;\Gamma}_{\alert{[\idS]_\Ty}} t
  \qquad\qquad
  [;]_{\Tm}    : \implicit{\Gamma, \Delta, \Theta, \sigma, \tau, t}\; {[\sigma ; \tau]\,t} =^{\Tm\;\Gamma}_{\alert{[;]_{\Ty}}}{[ \sigma ]\,[ \tau ]\,t}
\]
equivalent to the following (homogeneous) identities
\[
  \transfib{\Tm\;\Gamma}{[\idS]_{\Ty}}{[\idS]\,t} =^{\Tm\,\Gamma\,A} t
  \quad\text{and}\quad
  \transfib{\Tm\;\Gamma}{[;]_{\Ty}}{[\sigma;\tau]\,t} =^{\Tm\,\Gamma\,([\sigma]\,[\tau]A)} [\sigma]\,[\tau]\,t
\]
respectively.
Constructors introduced so far complete the definition of substitution calculus.

The use of transports in the formal definition fixes type mismatches but hinders equational reasoning about these terms even for a simple fact below.
\begin{example}\label{ex:pi2-comp}
  Given substitutions $\sigma : \Sub\;\Gamma\; \Delta$ and $\tau : \Sub\;\Delta\;(\Theta, A)$ for any $A : \Ty\,\Theta$, we may apply the projection $\pi_2$ to the composite $(\sigma; \tau)$ to access the first term $\pi_2(\sigma; \tau)$ of type $[\sigma;\tau] A$ under the context $\Gamma$, and this term should be equal to the first term $\pi_2\,\tau$ of $\tau$ after applying the substitution $\sigma$. 
  In short, the following identity apparently holds
  \[
    \pi_2\,(\sigma ; \tau) = [\sigma] (\pi_2\,\tau)
  \]
  by a back-of-the-envelope calculation
  \begin{equation} \label{eq:pi2-comp-proof}
    \pi_2\,(\sigma ; \tau) 
    = \pi_2\,(\sigma; (\pi_1\,\tau, \pi_2\,\tau))
    = \pi_2\,(\sigma;\pi_1\,\tau, [\sigma]\,(\pi_2\,\tau))
    = [\sigma] (\pi_2\,\tau).
  \end{equation}
  Yet, the left-hand side is a term of type $[\pi_1\,(\sigma;\tau)] A$, but the other is $[\sigma] [\pi_1\,\tau] A$.
  Hence the above identity does not even make sense, since their types do not match.
  Alas, instead, we have to write $\pi_2\,(\sigma ; \tau) =^{\Tm\,\Gamma}_{p} [\sigma] (\pi_2\,\tau)$ for some $p: [\pi_1(\sigma; \tau)]\,A = [\sigma]\,[\pi_1\tau]\,A$ or, equivalently
  \begin{equation}\label{eq:pi2-comp-real-proof}
    ([\pi_1(\sigma; \tau)]\,A, \pi_2(\sigma; \tau)) =^{(A : \Ty\,\Gamma) \times (\Tm \Gamma A)} ([\sigma]\,[\pi_1\,\tau]A, [\sigma] (\pi_2\,\tau))
  \end{equation}
  as inhabitants of a $\Sigma$-type, so we can reason about term equalities along with type equalities.
  Moreover, in~\eqref{eq:pi2-comp-proof} we have used the rule $\mathsf{concat}$ which introduced another transported term, so we will have to remove that $\transp$ to derive the right hand side.

  To better illustrate the annoyance, note that a complete proof of \eqref{eq:pi2-comp-real-proof} requires us to show each of following equations:
  \begin{alignat*}{3}
         & ([\pi_1(\sigma; \tau)]\,A                    &&, \pi_2\,(\sigma ; \tau)) \\
    = {} & ([\pi_1(\sigma; (\pi_1\tau , \pi_2\tau))]\,A &&, \pi_2\,(\sigma; (\pi_1\,\tau, \pi_2\,\tau))) \\
    = {} & ([\pi_1(\sigma;\pi_1\,\tau, \transfib{}{[;]_{\Ty}^{-1}}{\color{black}[ \sigma ] (\pi_2\;\tau)})]\,A &&, \pi_2\,(\sigma;\pi_1\,\tau, \transfib{}{[;]_{\Ty}^{-1}}{\color{black}[ \sigma ] (\pi_2\;\tau)})) \\
    = {} & ([\sigma;\pi_1\,\tau]\,A &&, \transfib{}{[;]_{\Ty}^{-1}}{\color{black}[ \sigma ] (\pi_2\;\tau)})) \nonumber \\
    = {} & (\sub{\sigma}{\sub{\pi_1\tau}{A}} &&, [\sigma] (\pi_2\,\tau)). \nonumber
  \end{alignat*}
  The first three equations corresponding to \eqref{eq:pi2-comp-proof} follow with Lemma~2.3.4 in \cite{UFP2013}, while the last has nothing to do with substitution calculus but the path lifting property (\cite[Lemma~2.3.2]{UFP2013}).
\end{example}

@@

One possibility is to define the substitution calculus as a quotient inductive-inductive-recursive type, where type substitution is defined by recursion, and hence computes.
%Now that $[ \idS ]_{\Ty}\;A = A$ and $[ \sigma ; \tau ]_{\Ty}\;A = [ \sigma ]_{\Ty}\;([ \tau ]_{\Ty}\;A)$ do not make $[\blank]_{\Ty}\blank$ a (total) function, we may be tempted to define it by
%\begin{alignat*}{3}
%[\blank]_{\Ty} \blank &: \implicit{\Gamma, \Delta} \; \Sub\;\Gamma\;\Delta \to \Ty\;\Delta \to \Ty\;\Gamma \\
%[ \sigma ]_{\Ty}\;\UU & = \UU
%\end{alignat*}
Note that this does not follow the schema of inductive-recursive definitions given by Dybjer and Setzer~\cite{Dybjer2000,Dybjer2003}, since the codomain of the recursive function is the inductively defined typed itself.\footnote{%
  While the general schema for this more liberal notion of induction-recursion remains underdeveloped, this definition is accepted in \Agda and used in other formalisations of type theory~\cite{Danielsson2006}.}
%Nevertheless, $[\idS]_{\Ty}\,A$ and $[\sigma; \tau]_{\Ty}\,A$ do not reduce to~$A$ and~$[\sigma]_{\Ty}\,[\tau]_{\Ty}\,A$ respectively, even though they are provable.
%This definition fails to remove any transport in the definition of substitution calculus.
%
Since we want type substitution to both act structurally on types and commute with identities and composition, we could try to define it recursively as follows:
\begin{alignat}{3}
[\blank]_{\Ty} \blank &: \implicit{\Gamma, \Delta} \; \Sub\;\Gamma\;\Delta \to \Ty\;\Delta \to \Ty\;\Gamma \nonumber \\
[ \sigma ]_{\Ty}       \;\UU  & = \UU                                   \label{eq:type-sub-at2-1} \\
[ \idS ]_{\Ty}         \;A    & = A                                     \label{eq:type-sub-at2-2} \\
[ \sigma ; \tau ]_{\Ty}\;A    & = [ \sigma ]_{\Ty}\;([ \tau ]_{\Ty}\;A) \label{eq:type-sub-at2-3}
\end{alignat}
As explained in \cref{sec:meta:overlapping}, this definition does \emph{not} reduce \eqref{eq:type-sub-at2-2} or \eqref{eq:type-sub-at2-3} \emph{strictly}, and \eqref{eq:type-sub-at2-2} or \eqref{eq:type-sub-at2-3} will fire only if we know that $A$ is not $\UU$.
%In the above definition, only the eliminator for $\Ty\;\Delta$ will be considered using the first-match semantics, which amounts to the following definition:
%\begin{alignat*}{3}
%[\blank]_{\Ty} \blank &: \implicit{\Gamma, \Delta} \; \Sub\;\Gamma\;\Delta \to \Ty\;\Delta \to \Ty\;\Gamma \\
%[ \sigma ]_{\Ty}\;\UU  & \reduce \UU
%\end{alignat*}
%where $\reduce$ is used to emphasise the reduction (which entails the strict equality): we have done nothing different from our first attempt! 
%As a result, neither  are strict.
We may swap \eqref{eq:type-sub-at2-1} with \eqref{eq:type-sub-at2-2} and \eqref{eq:type-sub-at2-3} to make \eqref{eq:type-sub-at2-2} and \eqref{eq:type-sub-at2-3} strict, but then \eqref{eq:type-sub-at2-1} will be translated to 
\[
[ \emptysub ]_{\Ty}\;\UU         \reduce \UU, \qquad
[ \sigma, t ]_{\Ty}\;\UU         \reduce \UU, \qquad\text{and}\qquad 
[ \pi_1\,\sigma ]_{\Ty}\;\UU     \reduce \UU,
\]
where we have written $t \reduce u$ for $t \equiv u$ to emphasise the reduction behaviour. Thus in this case we have again lost the strict identity $[\sigma]\;\UU \equiv \UU$.
%
%making $[\sigma]\;\UU = \UU$ propositional instead.
%Not being able to reduce $[ \sigma ]_{\Ty}\;\UU$ to $\UU$ would make our later extension with a type $\El\,a$ for $a : \Tm\;\Gamma\;\UU$ require another transport in the definition.
%Likewise, for any extension with a type former such as $\Pi$-types, structural rules would still be propositional.
%That is, using the standard semantics of function definition, we are not able to make type substitution strict for \emph{both} $\Sub\;\Gamma\;\Delta$ and $\Ty\;\Delta$ at the same time.

@@

\subsection{Adding a universe} \label{subsec:SC+U}
In this section, we extend substitution calculus with an empty universe $\UU$ of small types~$\El\,u$.
For a QII definition \cite{Altenkirch2016a}, this extension adds the following constructors
\begin{alignat*}{3}
  \El   & : \implicit{\Gamma} && \Tm\,\Gamma\;\UU \to \Ty\,\Gamma \\
  []\El & : \implicit{\Gamma, \Delta, \sigma, u} && [ \sigma ]\,(\El\, u) =^{\Tm\;\Gamma\;\UU} \El\,(\alert{\transfib{\Tm\;\Gamma}{[]\UU}{\color{black}[\sigma]{u}}})
\end{alignat*}
%as the type $\UU$ of universe and its structural rule $[]\UU$ have been added to avoid a trivial inductive type.
The type substitution $[\sigma](\El\,u)$ is, informally, the type of the term substitution $[\sigma]\,u$ for $u : \Tm\;\Delta\;\UU$, and the transport is needed to coerce $[\sigma]\,u : \Tm\;\Gamma\;([\sigma]\;\UU)$ into $\Tm\;\Gamma\;\UU$.

@@

\subsection{Adding other type formers} \label{subsec:SC+U+Pi+more}
\LT{%
\begin{enumerate}
  \item Discuss the extensional identity type.
  \item Discuss Coquand universes.
  \item Discuss the unit type.
  \item Discuss the Boolean type.
  \item Discuss the type of natural numbers.
\end{enumerate}
}
\Fredrik{I think we should keep this brief. I started doing the natural numbers in Agda, but there were some problems with the rewrite $[\sigma] \mathsf{Nat} = \mathsf{Nat}$ not firing in the type of the eliminator...}

\section{The elimination principle and models} \label{sec:models}
\LT{%
\begin{enumerate}
  \item Show that two definitions are equivalent (ongoing).
  \item Show that methods are equivalent (ongoing).
\end{enumerate}
}
With the QIIR definition of a type theory, we show how $\Ctx$, $\Ty\,\Gamma$, $\Sub\;\Gamma\;\Delta$, and $\Tm\,\Gamma\,A$ are eliminated to another types by functions subject to equations corresponding to point constructors, equality constructors, and function clauses (\cref{subsec:eliminator}).
Then, we can give this QIIR definition of type theory a standard model (\cref{subsec:std-model}) and compare our QIIR definition with its QII counterpart (\cref{subsec:equivalence}).

@@

\begin{remark} \label{re:method-recursion-IIR}
Note again that our definition is not an instance of Dybjer and Setzer's schema of inductive-recursion where simultaneously defined functions do not require corresponding interpretation, as they target meta-theoretic types instead of inductive types being defined.
\end{remark}

>>>>

We work in intensional Martin-L\"of type theory with uniqueness of identity proofs (\emph{UIP} for short). In \cref{sec:models,sec:formal-implementation}, we also assume function extensionality and $\eta$-rules for $\Pi$-types and $\Sigma$-types.
%
We mostly follow the notational conventions of the HoTT book~\cite{UFP2013} (even if we do not assume the univalence axiom), except that we write dependent function types using Agda notation $(x : A) \to B\,x$.
%
We are further inspired by Agda in writing $\implicit{x}\;B\,x$ for $(x : A) \to B\,x$ when $x$ is treated as an `implicit' argument which is not given explicitly, but instead can be figured out from context.
We write $\Set_i$ for the cumulative hierarchy of universes with the level index $i$ suppressed.

@@

Quotient inductive types (QITs) are a special case of higher inductive types~\cite{Lumsdaine2020} from homotopy type theory, where constructors can not only introduce new elements, but also new equalities between the introduced elements.
%
Since we assume UIP globally, all QITs are implicitly truncated to be sets in the sense of homotopy type theory.
%
For example, the type of \emph{join lists} $\JList{A}$ can be given by the QIT
with the following constructors:
\begin{alignat*}{3}
  \Jnil                    & : \implicit{A}      && \JList{A} \\
  \Jsing{\blank}           & : \implicit{A}      && A \to \JList{A} \\
  \Jconcat{\blank}{\blank} & : \implicit{A}      && \JList{A} \to \JList{A} \to \JList{A} \\
  \mathsf{idl}             & : \implicit{\ell}\; && (\Jconcat{\ell}{\Jnil}) =^{\JList{A}} \ell \\
  \mathsf{idr}             & : \implicit{\ell}\; && (\Jconcat{\Jnil}{\ell}) =^{\JList{A}} \ell \\
  \mathsf{assoc}           & : \implicit{\ell_0, \ell_1, \ell_2}\; && \Jconcat{(\Jconcat{\ell_0}{\ell_1})}{\ell_2} =^{\JList{A}} \Jconcat{\ell_0}{(\Jconcat{\ell_1}{\ell_2})}
\end{alignat*}

@@

\subsection{Definitions by overlapping patterns} \label{sec:meta:overlapping}
\begin{figure} 
\begin{minipage}[t]{.48\textwidth}
  \begin{alignat*}{5}
    \blank  & + \blank     && : \N \to \N \to \N \\
    0       &  + m         && = m  \\
  n         &  + 0         && = m  \\
  (\suc\;n) &  + m         && = \suc\;(n + m) \\
  n         &  + (\suc\;m) && = \suc\;(n + m)
\end{alignat*}
\end{minipage}
\begin{minipage}[t]{.5\textwidth}
\begin{alignat*}{5}
  \blank         & +'\blank     && : \N \to \N \to \N \\
  0              & +' m         && \defeq m  \\
  (\suc\;n)      & +' 0         && \defeq \suc\;n \\
  (\suc\;n)      & +' (\suc\;m) && \defeq \suc\;(n +' \suc\;m)
\end{alignat*}
\end{minipage}
\caption{Two seemingly different definitions of addition}
\label{fig:additions}
\end{figure}
Functions are typically defined informally using pattern matching for convenience and readability. Formally, however, they are either defined using an explicit pattern-matching construct or elaborated into case trees, which correspond to the \emph{eliminator}, following \emph{first-match} semantics~\cite{Cockx2020a} and employing termination and completeness checking.
This elaboration offers the conventional notation, while making computational behaviour opaque and favouring asymmetric definitions.
For example, two definitions of addition in \cref{fig:additions} are the same
after elaboration in \Agda: eliminating the first argument of $+$ distinguishes $0$ and $\suc\;n$ and refines the second and the third clauses into clauses of $+'$, making the fourth clause useless.
Each clause of $+'$ is strict (hence the symbol $\defeq$ is used) while clauses of $+$ are not necessarily strict by definition.
%This semantics for function definitions decides how definitions reduce and therefore how identities are proved.

Cockx~\cite{Cockx2014} explores an alternative semantics for pattern matching, where all clauses are strict, even in the presence of \emph{overlapping patterns}.
Thus, when a function $f$ is defined with overlapping patterns, the notation $f\;\overrightarrow{p_i} \reduce t_i$ is used to emphasize its reduction behaviour.
Under this semantics, each clause of $+$ is strictly equal to its right-hand side, allowing the symmetry addition to be proved commutative directly by induction on $n$ and $m$.

@@

The first and second projections of $(\sigma, t)$ should obviously be $\sigma$ and $t$ respectively
\begin{alignat*}{5}
  \pi_1\beta      & : \implicit{\Gamma, \Delta, \Theta, \sigma, A, t}\; && \pi_1(\sigma , t)        && =^{\Sub\,\Gamma,\Delta} &&  \sigma \\
  \pi_2\beta      & : \implicit{\Gamma, \Delta, \Theta, \sigma, A, t}\; && \pi_2(\sigma , t)        && =^{\Tm\,\Gamma\,([\blank]\,A)}_{\alert{\pi_1\beta}} &&  t, 
\end{alignat*}
but again, note that the type of $\pi_2\beta$ only makes sense because we can transport along $\pi_1\beta$ to make sure both sides of the equation have the same type: $\pi_2(\sigma, t)$ is an habitant of $\Tm\;\Gamma\;([\pi_1\,(\sigma, t)]\,A)$ instead of $\Tm\;\Gamma\;([\sigma]\,A)$.
%whereas $\pi_2(\sigma, t)$ is an habitant of $\Tm\;\Gamma\;([\pi_1\,(\sigma, t)]\,A)$ instead of $\Tm\;\Gamma\;([\sigma]\,A)$ on the right-hand side. 
%Therefore, we have to transport $\pi_2(\sigma, t)$ along $\pi_1\beta$, so the above equality constructor is, in fact, $\transfib{\Tm\;\Gamma\;([\blank]\,A)}{\pi_1\beta}{\pi_2(\sigma, t)} =^{\Tm\;\Gamma\;([\sigma]\,A)} t$.
%
Similarly, for term substitution, terms are transported along the corresponding rules:
\[
  [\idS]_{\Tm} : \implicit{\Gamma, A, t}\; {[\,\idS\,]\,t} =^{\Tm\;\Gamma}_{\alert{[\idS]_\Ty}} t
  \qquad\qquad
  [;]_{\Tm}    : \implicit{\Gamma, \Delta, \Theta, \sigma, \tau, t}\; {[\,\sigma ; \tau\,]\,t} =^{\Tm\;\Gamma}_{\alert{[;]_{\Ty}}}{[\, \sigma\,]\,[\,\tau\,]\,t}
\]
%equivalent to the following (homogeneous) identities
%\[
%  \transfib{\Tm\;\Gamma}{[\idS]_{\Ty}}{[\idS]\,t} =^{\Tm\,\Gamma\,A} t
%  \quad\text{and}\quad
%  \transfib{\Tm\;\Gamma}{[;]_{\Ty}}{[\sigma;\tau]\,t} =^{\Tm\,\Gamma\,([\sigma]\,[\tau]A)} [\sigma]\,[\tau]\,t
%\]
%respectively.
Constructors introduced so far complete the inductive definition of substitution calculus.

The use of transports in the formal definition fixes type mismatches but hinders equational reasoning about these terms even for a simple fact below.
\begin{example}\label{ex:pi2-comp}
  Given substitutions $\sigma : \Sub\;\Gamma\; \Delta$ and $\tau : \Sub\;\Delta\;(\Theta, A)$ for any $A : \Ty\,\Theta$, we may apply the projection $\pi_2$ to the composite $(\sigma; \tau)$ to access the first term $\pi_2(\sigma; \tau)$ of type $[\sigma;\tau] A$ under the context $\Gamma$, and this term should be equal to the first term $\pi_2\,\tau$ of $\tau$ after applying the substitution $\sigma$. 
  In short, the following identity apparently holds
  \[
    \pi_2\,(\sigma ; \tau) = [\sigma] (\pi_2\,\tau)
  \]
  and is checked by a back-of-the-envelope calculation
  \begin{equation} \label{eq:pi2-comp-proof}
    \pi_2\,(\sigma ; \tau) 
    = \pi_2\,(\sigma; (\pi_1\,\tau, \pi_2\,\tau))
    = \pi_2\,(\sigma;\pi_1\,\tau, [\sigma]\,(\pi_2\,\tau))
    = [\sigma] (\pi_2\,\tau).
  \end{equation}
  Yet, the left-hand side is a term of type $[\pi_1\,(\sigma;\tau)] A$, but the other is $[\sigma] [\pi_1\,\tau] A$.
  Hence the above identity does not even make sense, since their types do not match.
  Alas, instead, we have to write $\pi_2\,(\sigma ; \tau) =^{\Tm\,\Gamma}_{p} [\sigma] (\pi_2\,\tau)$ for some $p: [\pi_1(\sigma; \tau)]\,A = [\sigma]\,[\pi_1\tau]\,A$ or, equivalently
  \begin{equation}\label{eq:pi2-comp-real-proof}
    ([\pi_1(\sigma; \tau)]\,A, \pi_2(\sigma; \tau)) =^{(A : \Ty\,\Gamma) \times (\Tm \Gamma A)} ([\sigma]\,[\pi_1\,\tau]A, [\sigma] (\pi_2\,\tau))
  \end{equation}
  as inhabitants of a $\Sigma$-type, so we can reason about term equalities along with type equalities.
  Moreover, in~\eqref{eq:pi2-comp-proof} we have used the rule $\mathsf{concat}$ which introduced another transported term, so we will have to remove that $\transp$ to derive the right hand side.

  To better illustrate the annoyance, note that \eqref{eq:pi2-comp-real-proof} can be proved by following equations:
  \begin{alignat*}{3}
         & ([\pi_1(\sigma; \tau)]\,A                    &&, \pi_2\,(\sigma ; \tau)) \\
    = {} & ([\pi_1(\sigma; (\pi_1\tau , \pi_2\tau))]\,A &&, \pi_2\,(\sigma; (\pi_1\,\tau, \pi_2\,\tau))) \\
    = {} & ([\pi_1(\sigma;\pi_1\,\tau, \transfib{}{[;]_{\Ty}^{-1}}{\color{black}[ \sigma ] (\pi_2\;\tau)})]\,A &&, \pi_2\,(\sigma;\pi_1\,\tau, \transfib{}{[;]_{\Ty}^{-1}}{\color{black}[ \sigma ] (\pi_2\;\tau)})) \\
    = {} & ([\sigma;\pi_1\,\tau]\,A &&, \transfib{}{[;]_{\Ty}^{-1}}{\color{black}[ \sigma ] (\pi_2\;\tau)})) \nonumber \\
    = {} & (\sub{\sigma}{\sub{\pi_1\tau}{A}} &&, [\sigma] (\pi_2\,\tau)). \nonumber
  \end{alignat*}
  The first three equations corresponding to \eqref{eq:pi2-comp-proof} follow with Lemma~2.3.4 in \cite{UFP2013}, while the last has nothing to do with substitution calculus but the path lifting property (\cite[Lemma~2.3.2]{UFP2013}).
\end{example}

@@

One possibility is to define the substitution calculus as a quotient inductive-inductive-recursive type, where type substitution is defined by recursion, and hence computes.
%Now that $[ \idS ]_{\Ty}\;A = A$ and $[ \sigma ; \tau ]_{\Ty}\;A = [ \sigma ]_{\Ty}\;([ \tau ]_{\Ty}\;A)$ do not make $[\blank]_{\Ty}\blank$ a (total) function, we may be tempted to define it by
%\begin{alignat*}{3}
%[\blank]_{\Ty} \blank &: \implicit{\Gamma, \Delta} \; \Sub\;\Gamma\;\Delta \to \Ty\;\Delta \to \Ty\;\Gamma \\
%[ \sigma ]_{\Ty}\;\UU & = \UU
%\end{alignat*}
Note that this does not follow the schema of inductive-recursive definitions given by Dybjer and Setzer~\cite{Dybjer2000,Dybjer2003}, since the codomain of the recursive function is the inductively defined typed itself.\footnote{%
  While the general schema for this more liberal notion of induction-recursion remains underdeveloped, this definition is accepted in \Agda and used in other formalisations of type theory~\cite{Danielsson2006}.}
%Nevertheless, $[\idS]_{\Ty}\,A$ and $[\sigma; \tau]_{\Ty}\,A$ do not reduce to~$A$ and~$[\sigma]_{\Ty}\,[\tau]_{\Ty}\,A$ respectively, even though they are provable.
%This definition fails to remove any transport in the definition of substitution calculus.
%
Since we want type substitution to both act structurally on types and commute with identities and composition, we could try to define it recursively as follows:
\begin{alignat}{3}
[\blank]_{\Ty} \blank &: \implicit{\Gamma, \Delta} \; \Sub\;\Gamma\;\Delta \to \Ty\;\Delta \to \Ty\;\Gamma \nonumber \\
[ \sigma ]_{\Ty}       \;\UU  & = \UU                                   \label{eq:type-sub-at2-1} \\
[ \idS ]_{\Ty}         \;A    & = A                                     \label{eq:type-sub-at2-2} \\
[ \sigma ; \tau ]_{\Ty}\;A    & = [ \sigma ]_{\Ty}\;([ \tau ]_{\Ty}\;A) \label{eq:type-sub-at2-3}
\end{alignat}
As explained in \cref{sec:meta:overlapping}, this definition does \emph{not} reduce \eqref{eq:type-sub-at2-2} or \eqref{eq:type-sub-at2-3} \emph{strictly}, and \eqref{eq:type-sub-at2-2} or \eqref{eq:type-sub-at2-3} will fire only if we know that $A$ is not $\UU$.
%In the above definition, only the eliminator for $\Ty\;\Delta$ will be considered using the first-match semantics, which amounts to the following definition:
%\begin{alignat*}{3}
%[\blank]_{\Ty} \blank &: \implicit{\Gamma, \Delta} \; \Sub\;\Gamma\;\Delta \to \Ty\;\Delta \to \Ty\;\Gamma \\
%[ \sigma ]_{\Ty}\;\UU  & \reduce \UU
%\end{alignat*}
%where $\reduce$ is used to emphasise the reduction (which entails the strict equality): we have done nothing different from our first attempt! 
%As a result, neither  are strict.
We may swap \eqref{eq:type-sub-at2-1} with \eqref{eq:type-sub-at2-2} and \eqref{eq:type-sub-at2-3} to make \eqref{eq:type-sub-at2-2} and \eqref{eq:type-sub-at2-3} strict, but then \eqref{eq:type-sub-at2-1} will be elaborated to 
\[
[ \emptysub ]_{\Ty}\;\UU         \judgeq \UU, \qquad
[ \sigma, t ]_{\Ty}\;\UU         \judgeq \UU, \qquad\text{and}\qquad 
[ \pi_1\,\sigma ]_{\Ty}\;\UU     \judgeq \UU,
\]
Thus in this case we have again lost the desired strict identity $[\sigma]\;\UU \equiv \UU$.
%
%making $[\sigma]\;\UU = \UU$ propositional instead.
%Not being able to reduce $[ \sigma ]_{\Ty}\;\UU$ to $\UU$ would make our later extension with a type $\El\,a$ for $a : \Tm\;\Gamma\;\UU$ require another transport in the definition.
%Likewise, for any extension with a type former such as $\Pi$-types, structural rules would still be propositional.
%That is, using the standard semantics of function definition, we are not able to make type substitution strict for \emph{both} $\Sub\;\Gamma\;\Delta$ and $\Ty\;\Delta$ at the same time.

@@

\subsection{Adding a universe} \label{subsec:SC+U}
Now we extend substitution calculus with an empty universe $\UU$ of small types~$\El\,u$.
For a QII definition \cite{Altenkirch2016a}, this extension adds the following constructors
\begin{alignat*}{3}
  \El   & : \implicit{\Gamma} && \Tm\,\Gamma\;\UU \to \Ty\,\Gamma \\
  []\El & : \implicit{\Gamma, \Delta, \sigma, u} && [ \sigma ]\,(\El\, u) =^{\Tm\;\Gamma\;\UU} \El\,(\alert{\transfib{\Tm\;\Gamma}{[]\UU}{\color{black}[\sigma]{u}}})
\end{alignat*}
%as the type $\UU$ of universe and its structural rule $[]\UU$ have been added to avoid a trivial inductive type.
The type substitution $[\sigma](\El\,u)$ is, informally, the type of the term substitution $[\sigma]\,u$ for $u : \Tm\;\Delta\;\UU$, and the transport is needed to coerce $[\sigma]\,u : \Tm\;\Gamma\;([\sigma]\;\UU)$ into $\Tm\;\Gamma\;\UU$.

@@

\subsection{Adding other type formers} \label{subsec:SC+U+Pi+more}
\LT{%
\begin{enumerate}
  \item Discuss the extensional identity type.
  \item Discuss Coquand universes.
  \item Discuss the unit type.
  \item Discuss the Boolean type.
%  \item Discuss the type of natural numbers.
\end{enumerate}
}
\Fredrik{I think we should keep this brief. I started doing the natural numbers in Agda, but there were some problems with the rewrite $[\sigma] \mathsf{Nat} = \mathsf{Nat}$ not firing in the type of the eliminator...}

\section{The elimination principle and models} \label{sec:models}

With the QIIR definition of a type theory, we show how $\Ctx$, $\Ty\,\Gamma$, $\Sub\;\Gamma\;\Delta$, and $\Tm\,\Gamma\,A$ are eliminated to another types by functions subject to equations corresponding to point constructors, equality constructors, and function clauses (\cref{subsec:eliminator}).
Then, we can give this QIIR definition of type theory a standard model (\cref{subsec:std-model}) and compare our QIIR definition with its QII counterpart (\cref{subsec:equivalence}).

@@

\begin{remark} \label{re:method-recursion-IIR}
Note that a simultaneously defined function does not need a corresponding interpretation for Dybjer and Setzer's inductive-recursive types, as their framework targets meta-theoretic types rather than the inductive types being defined.
\end{remark}
